---
title: "Blue Space Analysis"
author: "Matthew Morgan"
date: "2024-03-11"
output: html_document
---

```{r load all required packages}
library(readr)
library(dplyr)
library(ggplot2)
library(vegan)
library(tidyverse)
library(psych)
library(MASS)
library(car)
library(lme4)
library(mgcv)
library(influence.ME)
library(ggpubr)
library(gridExtra)
library(gratia)
library(broom)
library(corrplot)
library(ggcorrplot)
```

```{r set_directory}
setwd("C:/users/721049/OneDrive - hull.ac.uk/GitHub_Cities_Analysis")
getwd()
```

```{r read in data}
df_1 <- read_csv("UKCEHLCM_final_df_200.csv", show_col_types = FALSE)
```

```{r create new columns}
column_names <- colnames(df_1)
print(column_names)

# new columns in df for looking at blue space

df_2 <- df_1 %>%
  mutate(
    built_cover = rowSums(across(c(Suburban, Urban)), na.rm = TRUE),
    grassland = rowSums(across(c(`Improved grassland`, `Neutral grassland`, `Calacareous grassland`, `Acid grassland`)), na.rm = TRUE),
    tree_cover = rowSums(across(c(`Deciduous woodland`, `Coniferous woodland`)), na.rm = TRUE),
    coastal_bg = rowSums(across(c(Saltwater, `Supralittoral rock`, `Supralittoral sediment`, `Littoral rock`, `Littoral sediment`, Saltmarsh, foreshore_area, tidal_water_area)), na.rm = TRUE),
    inland_bg = rowSums(across(c(surface_water_area, Freshwater, Fen, Bog)), na.rm = TRUE),
    other_cover = rowSums(across(c(`Heather grassland`, `Heather`, `Inland rock`)), na.rm = TRUE),
    lit_sediment_foreshore = rowSums(across(c(`Littoral sediment`, foreshore_area)), na.rm = TRUE),
    saltwater_tidalwater = rowSums(across(c(Saltwater, tidal_water_area)), na.rm = TRUE),
    freshwater_surfacewater = rowSums(across(c(Freshwater, surface_water_area)), na.rm = TRUE),
  ) %>%
  mutate(
    greenspace = rowSums(across(c(grassland, tree_cover, other_cover, Arable)), na.rm = TRUE),
    bluespace = rowSums(across(c(inland_bg, coastal_bg)), na.rm = TRUE),
    all_inland = rowSums(across(c(Freshwater, surface_water_area)), na.rm = TRUE),
    all_tidal = rowSums(across(c(Saltwater, tidal_water_area)), na.rm = TRUE),
    check = rowSums(across(c(built_cover, greenspace, bluespace)), na.rm = TRUE)
  )

buffered_city <- (df_2[1, 28])
classified_land <- (df_2[1, 35]) 
buffered_city
classified_land
```

Note that land use data has been absorbed into land cover by this point, 
so there are 21 land cover types, as in the original UKCEH LC map.
```{r calculating landcover diversity indicies}

all_cover <- c('Deciduous woodland', 'Coniferous woodland', 'Arable', 'Improved grassland',
                 'Neutral grassland', 'Calacareous grassland', 'Acid grassland', 'Fen', 
                 'Heather', 'Heather grassland', 'Bog', 'Inland rock', 'saltwater_tidalwater', 
                 'freshwater_surfacewater', 'Supralittoral rock', 'Supralittoral sediment', 
                 'Littoral rock', 'lit_sediment_foreshore', 'Saltmarsh', 'Urban', 
                 'Suburban')

# Check no land cover types are missing from the all_cover vector by cross checking

df_3 <- df_2 %>%
  rowwise() %>%
  mutate(div.check = sum(c_across(all_of(all_cover)))) %>%
  ungroup()
print(df_2$check[1]) # 22267785
print(df_3$div.check[1]) # 22267785

# Create diversity columns

df_2 <- df_2 %>%
  rowwise() %>%
  mutate(
    shannon = diversity(c_across(all_of(all_cover)), index = "shannon"),
    simpsons = diversity(c_across(all_of(all_cover)), index = "simpson")
  ) %>%
  ungroup()
```

```{r check zero influence}
# isolate one bua at a time
Aberdare <- filter(df_2, name == "Aberdare")

# Remove columns where all values are zero
Aberdare_clean <- Aberdare %>% select_if(~any(. != 0))
str(Aberdare_clean)

# Get the column names from the Aberdare_clean dataframe
aber_clean_names <- names(Aberdare_clean)

# Find what is in all_land_cover but not in Aberdare_clean
missing_in_aberclean <- setdiff(all_cover, aber_clean_names)

# Find what columns are in Aberdare_clean but not in all_land_cover
extra_in_aberclean <- setdiff(aber_clean_names, all_cover)

# Print the results
print(paste("Missing in Aberdare_clean:", toString(missing_in_aberclean))) # remove land cover that is not present
print(paste("Extra in Aberdare_clean:", toString(extra_in_aberclean))) 

# Aberdare cover
aberdare_cover <- c('Deciduous woodland', 'Coniferous woodland', 'Arable', 'Improved grassland',
                 'Neutral grassland', 'Calacareous grassland', 'Acid grassland', 'Fen', 
                 'Heather', 'Heather grassland', 'Bog', 'Inland rock', 
                 'freshwater_surfacewater', 'Urban', 
                 'Suburban')

Aberdare_clean_2 <- Aberdare_clean %>%
  rowwise() %>%
  mutate(
    shannon1 = diversity(c_across(all_of(aberdare_cover)), index = "shannon"),
    simpsons1 = diversity(c_across(all_of(aberdare_cover)), index = "simpson")
  ) %>%
  ungroup()

# No difference with or without zero's when calculating diversity of land cover. 
shan_with_zero <- (df_2[1, 50]) #1.48
simp_with_zero <- (df_2[1, 51]) #0.69
shan_without_zero <- (Aberdare_clean_2[1, 40]) # 1.48
simp_without_zero <- (Aberdare_clean_2[1, 41]) # 0.69
t1 <- shan_with_zero - shan_without_zero 
t2 <- simp_with_zero - simp_without_zero
t1
t2
```

Check that the new columns are correct (should sum to the same value as lu_lc_total) 
```{r quality_checks}
colnames(df_2)
# 
check1 <- df_2[1:500, 35] # original lc_lu sum 
check2 <- df_2[1:500, 49] # check sum of bluespace, greenspace and greyspace

# Calculate the difference (all zero)
difference <- check1 - check2
formatted_difference <- round(difference, 3)
print(formatted_difference)

# Percentages of blue, green, grey
df_3 <- df_2 %>%
  mutate(
    rel_built = (built_cover / lc_lu_total) * 100,
    total_green = (greenspace / lc_lu_total) * 100,
    total_blue = (bluespace / lc_lu_total) * 100,
    )
```

```{r remove minor urban areas}
# remove sites with minor populations
removed_entries <- df_3[df_3$population_counts < 5000, ]

# Step 1: Remove population_counts >= 5000 (minor populations not suitable for this study)
df_3 <- df_3[df_3$population_counts >= 5000, ]
removed_entries

# export dataframe for other analysis (optional)
# export df
# write.csv(df_3, "analysis_5.csv", row.names = FALSE)
```

```{r descriptive_stats}
# Create summaries for green, blue and grey space
summary <- df_3 %>%
  summarise(across(c(rel_built, total_green, total_blue), 
                   list(mean = mean, sd = sd, min = min, max = max), na.rm = TRUE))
print(summary)

# Descriptive statistics
descriptive <- describe(df_3)
```

```{r blue space diagnostics}
# Subset
coastal <- df_3 %>% filter(class == "coastal") 
inland <- df_3 %>% filter(class == "inland") 
estuarine <- df_3 %>% filter(class == "estuarine") 

# Function to create histograms and Q-Q plots
create_plots <- function(data, title_prefix, color) {
  hist(data, main = paste("Histogram of", title_prefix, "Blue Space"), xlab = "Blue Space %", col = color, breaks = 20, cex.main = 2, cex.lab = 1.8, cex.axis = 1.8)
  qqnorm(data, main = paste("Q-Q Plot for", title_prefix, "Blue Space"), cex.main = 2, cex.lab = 1.8, cex.axis = 1.8)
  qqline(data, col = color)
}

# 
png("blue_diagnostic_plots_v3.png", width = 1500, height = 1000)

# canvas
par(mfrow = c(3, 2), mar = c(5, 5, 4, 2)) 

# plot
create_plots(coastal$total_blue, "Coastal", "blue")
create_plots(inland$total_blue, "Inland", "red")
create_plots(estuarine$total_blue, "Estuarine", "darkgreen")  # Corrected from inland to estuarine
```

```{r spearman's rank correlation - area}
# inland data vs bua size
inland_correlation_result <- cor.test(inland$total_blue, inland$original_bua_area, method = "spearman")
print(inland_correlation_result)  #  *inland data has a weak positive relationship, which is stat. significant
plot(inland$original_bua_area, inland$total_blue)
     
# coastal blue space and bua size
coastal_correlation_result <- cor.test(coastal$total_blue, coastal$original_bua_area, method = "spearman")
print(coastal_correlation_result) # ** coastal data has a weak positive relationship, which is stat. significant
plot(coastal$original_bua_area, coastal$total_blue)

# estuarine blue space and bua size
coastal_correlation_result <- cor.test(estuarine$total_blue, estuarine$original_bua_area, method = "spearman")
print(coastal_correlation_result) # ** coastal data has a weak positive relationship, which is stat. significant
plot(estuarine$original_bua_area, estuarine$total_blue)
```
```{r spearman's rank correlation - population counts}
# inland data vs bua size
inland_pop_correlation_result <- cor.test(inland$total_blue, inland$population_counts, method = "spearman")
print(inland_pop_correlation_result)  #  *inland data has a weak positive relationship, which is stat. significant
plot(inland$population_counts, inland$total_blue)
     
# coastal blue space and bua size
coastal_pop_correlation_result <- cor.test(coastal$total_blue, coastal$population_counts, method = "spearman")
print(coastal_pop_correlation_result) # ** coastal data has a weak positive relationship, which is stat. significant
plot(coastal$population_counts, coastal$total_blue)

# estuarine blue space and bua size
estuarine_pop_correlation_result <- cor.test(estuarine$total_blue, estuarine$population_counts, method = "spearman")
print(estuarine_pop_correlation_result) # ** coastal data has a weak positive relationship, which is stat. significant
plot(estuarine$population_counts, estuarine$total_blue)
```

```{r spearman's rank correlation - grey space}
# inland data vs bua size
inland_grey_correlation_result <- cor.test(inland$total_blue, inland$rel_built, method = "spearman")
print(inland_grey_correlation_result)  #  *inland data has a weak positive relationship, which is stat. significant
plot(inland$rel_built, inland$total_blue)
     
# coastal blue space and bua size
coastal_grey_correlation_result <- cor.test(coastal$total_blue, coastal$rel_built, method = "spearman")
print(coastal_grey_correlation_result) # ** coastal data has a weak positive relationship, which is stat. significant
plot(coastal$rel_built, coastal$total_blue)

# estuarine blue space and bua size
estuarine_grey_correlation_result <- cor.test(estuarine$total_blue, estuarine$rel_built, method = "spearman")
print(estuarine_grey_correlation_result) # ** coastal data has a weak positive relationship, which is stat. significant
plot(estuarine$rel_built, estuarine$total_blue)
```
```{r spearman's rank correlation - green space}
# inland data vs bua size
inland_green_correlation_result <- cor.test(inland$total_blue, inland$total_green, method = "spearman")
print(inland_green_correlation_result)  #  *inland data has a weak positive relationship, which is stat. significant
plot(inland$total_green, inland$total_blue)
     
# coastal blue space and bua size
coastal_green_correlation_result <- cor.test(coastal$total_blue, coastal$total_green, method = "spearman")
print(coastal_green_correlation_result) # ** coastal data has a weak positive relationship, which is stat. significant
plot(coastal$total_green, coastal$total_blue)

# estuarine blue space and bua size
estuarine_green_correlation_result <- cor.test(estuarine$total_blue, estuarine$total_green, method = "spearman")
print(estuarine_green_correlation_result) # ** coastal data has a weak positive relationship, which is stat. significant
plot(estuarine$total_green, estuarine$total_blue)
```

```{r diagnostic plots bluespace}
# check blue space normality
#
png("histograms_bluespace.png", width = 1000, height = 1200)

# Set plotting parameters
par(mfrow = c(4, 2), mar = c(2, 2, 2, 1), cex.main =1.5)

# Plot your histograms with titles
hist(inland$total_blue, main = "1. Inland Blue Space")
hist(log(inland$total_blue), main = "2. Log Inland Blue")
hist(coastal$total_blue, main = "3. Coastal Blue Space")
hist(log(coastal$total_blue), main = "4. Log of Coastal Blue Space")
hist(estuarine$total_blue, main = "5. Estuarine Blue Space")
hist(log(estuarine$total_blue), main = "6. Log of Estuarine Blue Space")
hist(sqrt(estuarine$total_blue), main = "7. Square Root of Estuarine Blue Space")
hist(sqrt(coastal$total_blue), main = "8. Square Root of Coastal Blue Space")

# Close the PNG device
dev.off()
```

```{r diagnostic plots greenspace}
png("histograms_greenpace.png", width = 1000, height = 1200)

# Set plotting parameters
par(mfrow = c(4, 2), mar = c(2, 2, 2, 1), cex.main =1.5)

# Check green space normality
par(mfrow = c(3,2))
hist(inland$total_green, main = "1. Inland Green Space")
hist(log(inland$total_green), main = "2. Log of Inland Green Space")
hist(coastal$total_green, main = "3. Coastal Green Space")
hist(log(coastal$total_green), main = "4. Coastal Green Space")
hist(estuarine$total_green, main = "5. Estuarine Green Space")
hist(log(estuarine$total_green), main = "6. Estuarine Green Space")

# Close the PNG device
dev.off()
```

```{r diagnostic plots greyspace}
png("histograms_greyspace.png", width = 1000, height = 1200)

# Set plotting parameters
par(mfrow = c(4, 2), mar = c(2, 2, 2, 1), cex.main =1.5)

# Check green space normality
par(mfrow = c(3,2))
hist(inland$rel_built, main = "1. Inland Grey Space")
hist(log(inland$rel_built), main = "2. Log of Inland Grey Space")
hist(coastal$rel_built, main = "3. Coastal Grey Space")
hist(log(coastal$rel_built), main = "4. Log of Coastal Grey Space")
hist(estuarine$rel_built, main = "5. Estuarine Grey Space")
hist(log(estuarine$rel_built), main = "6. Estuarine Grey Space")

# Close the PNG device
dev.off()
```

```{r diagnostic plots pop. counts}
png("histograms_populationcounts.png", width = 1000, height = 1200)

# Set plotting parameters
par(mfrow = c(4, 2), mar = c(2, 2, 2, 1), cex.main =1.5)

# Check green space normality
par(mfrow = c(3,2))
hist(inland$population_counts, main = "1. Inland Population Counts")
hist(log(inland$population_counts), main = "2. Log of Inland Population Counts")
hist(coastal$population_counts, main = "3. Coastal Population Counts")
hist(log(coastal$population_counts), main = "4. Log of Coastal Population Counts")
hist(estuarine$population_counts, main = "5. Estuarine Population Counts")
hist(log(estuarine$population_counts), main = "6. Log of Estuarine Population Counts")

# Close the PNG device
dev.off()
```

```{r diagnostic plots area}
png("histograms_bua.png", width = 1000, height = 1200)

# Set plotting parameters
par(mfrow = c(4, 2), mar = c(2, 2, 2, 1), cex.main =1.5)

# Check green space normality
par(mfrow = c(3,2))
hist(inland$original_bua_area, main = "1. Inland BUA Size")
hist(log(inland$original_bua_area), main = "2. Log of Inland BUA Size")
hist(coastal$original_bua_area, main = "3. Coastal BUA Size")
hist(log(coastal$original_bua_area), main = "4. Log of Coastal BUA Size")
hist(estuarine$original_bua_area, main = "5. Estuarine Population BUA Size")
hist(log(estuarine$original_bua_area), main = "6. Log of Estuarine BUA Size")

# Close the PNG device
dev.off()
```

```{r scatter plots}
# Coastal
# setup the range and aspect ratio so it can be square
x_range <- range(log(coastal$original_bua_area), na.rm = TRUE)
y_range <- range(coastal$total_blue, na.rm = TRUE)
aspect_ratio <- diff(x_range) / diff(y_range)

# Create the plot
pc <- ggplot(coastal, aes(x = log(original_bua_area), y = total_blue)) +
  geom_point(color = "#6495ED", size = 3, alpha = 0.8, show.legend = FALSE) +
  labs(title = "Coastal **",
       subtitle = "Spearman's Rank: RHO -0.33, p-value = 0.005",
       x = "Log of Size (m²)",
       y = "Blue Space %") +
  theme_light() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    axis.title.x = element_text(size = 12, margin = margin(t = 10, b = 15 )),
    axis.title.y = element_text(size = 12, margin = margin(r = 10, l = 10)),
      axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
    legend.position = "none") + 
  coord_fixed(ratio = aspect_ratio)
pc

#
ggsave("coastal_size.png", plot = pc, width = 6, height = 6, dpi = 300)

# Inland
x_range <- range(log(inland$original_bua_area), na.rm = TRUE)
y_range <- range(log(inland$total_blue), na.rm = TRUE)
aspect_ratio <- diff(x_range) / diff(y_range)
# Create the scatter plot
pi <- ggplot(inland, aes(x = log(original_bua_area), y = log(total_blue))) +
 geom_point(color = "#FFA07A", size = 3, alpha = 0.6, show.legend = FALSE)  +
  labs(title = "Inland **",
       subtitle = "Spearman's Rank: RHO 0.15, p-value = 0.002",
       x = "Log of Size (m²)",
       y = "Log of Blue Space (%)") +
  theme_light() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    axis.title.x = element_text(size = 12, margin = margin(t = 10, b = 15 )),
    axis.title.y = element_text(size = 12, margin = margin(r = 10, l = 10)),
    legend.position = "none",
      axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12)) +
  coord_fixed(ratio = aspect_ratio)
pi

#
ggsave("inland_size.png", plot = pi, width = 6, height = 6, dpi = 300)

# Estuarine
x_range <- range(log(estuarine$original_bua_area), na.rm = TRUE)
y_range <- range(log(estuarine$total_blue), na.rm = TRUE)
aspect_ratio <- diff(x_range) / diff(y_range)
# Create the scatter plot
pe <- ggplot(estuarine, aes(x = log(original_bua_area), y = log(total_blue))) +
 geom_point(color = "#95D8B5", size = 3, alpha = 2, show.legend = FALSE)  + 
  labs(title = "Estuarine",
       subtitle = "Spearman's Rank: RHO -0.12, p-value = 0.51",
       x = "Log of Size (m²)",
       y = "Log of Blue Space (%)") +
  theme_light() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    axis.title.x = element_text(size = 12, margin = margin(t = 10, b = 15 )),
    axis.title.y = element_text(size = 12, margin = margin(r = 10, l = 10)),
      axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
    legend.position = "none") +
  coord_fixed(ratio = aspect_ratio)
pe

# combine plots

combined_plot <- grid.arrange(pc ,pi, pe, ncol = 3)

# Save the combined plot as a JPEG file
ggsave("combined_scatter_plot.png", plot = combined_plot, width = 13, height = 6, dpi = 300)
```

## Deprivation data

```{r sub-set england}
# assign england to df_filtered
df_filtered <- df_3[df_3$country == "england", ]

```

```{r diagnostics for deprivation variables}
# Save plots to a PNG file
png("model_variable_dist_v1.png", width = 6000, height = 2100, res = 300)

# Set up a 7x3 layout and adjust text size
par(mfrow = c(3,8), cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, mar = c(4, 4, 2, 1))

# List of variables to plot
variables <- c("total_green", "total_blue", "imdd_weight_av", "shannon", "rel_built", "original_bua_area", "simpsons", "population_counts")

# Function to log-transform variables if needed
log_transform_if_needed <- function(var_name, df) {
  if (var_name %in% c("total_blue", "original_bua_area")) {
    return(log(df[[var_name]] + 1))  # Adding 1 to avoid log(0)
  } else {
    return(df[[var_name]])
  }
}

# Plot histograms
invisible(lapply(variables, function(var) {
  hist(log_transform_if_needed(var, df_filtered),
       main = paste("Histogram of", gsub("_", " ", var)),
       xlab = gsub("_", " ", var),
       col = "skyblue", border = "black")
}))

# Create boxplots
invisible(lapply(variables, function(var) {
  boxplot(log_transform_if_needed(var, df_filtered),
          main = paste("Boxplot of", gsub("_", " ", var)),
          xlab = gsub("_", " ", var),
          col = "skyblue", border = "black")
}))

# Create Q-Q plots
invisible(lapply(variables, function(var) {
  qqnorm(log_transform_if_needed(var, df_filtered),
         main = paste("Q-Q Plot of", gsub("_", " ", var)))
  qqline(log_transform_if_needed(var, df_filtered), col = "red", lwd = 2)
}))

# Close the PNG device to save the file
dev.off()


```

```{r correlation matrix}
df_filtered

# Calculate descriptive statistics
descriptive_stats <- describe(df_filtered %>% dplyr::select(total_green, total_blue, imdd_weight_av, shannon, rel_built, simpsons))
print(descriptive_stats)

# Correlation matrix
correlation_matrix <- cor(df_filtered %>% dplyr::select(total_green, total_blue, imdd_weight_av, shannon, rel_built, simpsons, original_bua_area, population_counts, long), use = "complete.obs", method = "spearman")
print(correlation_matrix)

# Visualize correlation matrix
corr_plot <- ggcorrplot(correlation_matrix, lab = TRUE)
print(corr_plot)

# Plot of notable correlation observed with green space and rel built (this is not part of the analysis just a visualistion of the some of the variables)
plot(df_filtered$rel_built, log(df_filtered$total_blue))
plot(df_filtered$shannon, log(df_filtered$total_blue))
plot(df_filtered$rel_built, df_filtered$total_green)
#plot(df_filtered$rel_built, df_filtered$total_blue)
```

```{r correlation test}

# Save plot to a PNG file with adjusted width and height for higher resolution
png("model_plots_v1.png", width = 3600, height = 1350, res = 300)  # Adjusted dimensions for 300 DPI

# Set up a 1x3 layout for the plots
par(mfrow = c(1, 3), cex.main = 1.2, cex.lab = 1.3, cex.axis = 1.2, mar = c(5, 5, 4, 2))

# Plot 1: Original BUA Area vs. Population Counts
plot(df_3$original_bua_area, df_3$population_counts,
     main = "Original BUA Area (metre sq.) vs. Pop. Counts (R=0.85)",
     xlab = "Original BUA Area", 
     ylab = "Population Counts",
     pch = 16, col = "black")

# Add the regression line
lm1 <- lm(population_counts ~ original_bua_area, data = df_3)
abline(lm1, col = "red", lwd = 2)

# Plot 2: Total Green vs. Relative Built
plot(df_3$total_green, df_3$rel_built,
     main = "Total Green vs. Total Grey (R=0.82)",
     xlab = "Total Green %", 
     ylab = "Total Grey %",
     pch = 16, col = "black")

# Add the regression line
lm2 <- lm(total_green ~ rel_built, data = df_3)
abline(lm2, col = "red", lwd = 2)

# Plot 3: Shannon vs. Simpsons
plot(df_3$simpsons, df_3$shannon,
     main = "Simpsons vs. Shannon (R=0.84)",
     xlab = "Simpsons Index", 
     ylab = "Shannon Index",
     pch = 16, col = "black")

# Add the regression line
lm3 <- lm(shannon ~ simpsons, data = df_3)
abline(lm3, col = "red", lwd = 2)

# Close the PNG device
dev.off()


```

```{r checking_vif}
# Fit a lm to check for VIF (all predictor variables which will be used)
model1 <- lm(imdd_weight_av ~ total_green + total_blue +
             shannon + original_bua_area + long + lat, data = df_filtered)
# summary(model1)

model2 <- lm(imdd_weight_av ~ rel_built + total_blue +
             shannon + population_counts + long + lat, data = df_filtered)
# summary(model2)

# assign and show values
vif_values1 <- vif(model1)
vif_values2 <- vif(model2)
print(vif_values1)
print(vif_values2)

# All independent variables are within correlation, VIF limits and suitable for further analysis. 

```

```{r linearity exploration}
# Dependent IMDD
# green
plot(df_filtered$total_green, df_filtered$imdd_weight_av)
with(df_filtered, cor.test(total_green, imdd_weight_av))

# blue
plot((log(df_filtered$total_blue)), df_filtered$imdd_weight_av)
with(df_filtered, cor.test(total_blue, imdd_weight_av, method = "spearman"))

# shannon
plot(df_filtered$shannon, df_filtered$imdd_weight_av)
with(df_filtered, cor(shannon, imdd_weight_av))

# size
plot((log(df_filtered$original_bua_area)), df_filtered$imdd_weight_av)
with(df_filtered, cor(original_bua_area, imdd_weight_av))

# simpson
plot((log(df_filtered$simpsons)), df_filtered$imdd_weight_av)
with(df_filtered, cor(simpsons, imdd_weight_av))

# grey
plot(df_filtered$rel_built, df_filtered$imdd_weight_av)
with(df_filtered, cor(rel_built, imdd_weight_av))

# pop size
plot((log(df_filtered$population_counts)), df_filtered$imdd_weight_av)
with(df_filtered, cor(population_counts, imdd_weight_av))
```
## General Additive Models 

After considering models types and examining the variables in the data set, a General Additive Model (GAM) was selected. This model was chosen as the dependent variable IMDD (Index of Mulitple Deprivation Decile) is normally distributed continuous data. However, the independent variables are combination of parametric and non-parametric data sets with varying degrees of linearity. GAM provides smoothing splines for non-linear variables and will incorporate categorical variables, making them suitable for mixed data (continuous and categorical variables). Various model permutations were compared to find the most explanatory, using variance explained, AIC and BIC.

```{r base model}
# fix naming error

# Rename 'long' to temporary placeholder
names(df_filtered)[names(df_filtered) == "long"] <- "temp_lat"

# Rename 'lat' to 'long'
names(df_filtered)[names(df_filtered) == "lat"] <- "long"

# Rename placeholder to 'lat'
names(df_filtered)[names(df_filtered) == "temp_lat"] <- "lat"


df_filtered$class <- as.factor(df_filtered$class)

gam_model1 <- gam(imdd_weight_av ~ total_green + s(log(total_blue)) + shannon + s(log(original_bua_area)) + class + lat + long, data = df_filtered, family = gaussian())

#
summary(gam_model1)
gam.check(gam_model1)
AIC(gam_model1)
```

base model - class and original bua
```{r gam model two}
gam_model2 <- gam(imdd_weight_av ~ total_green + s(log(total_blue)) + shannon + long, data = df_filtered, family = gaussian())

#
summary(gam_model2)
gam.check(gam_model2)
```
smooth all existing independent variables in the model 
```{r gam model three}
# Predicted variables
par(mfrow = c(2,2))
hist(df_filtered$shannon)
hist(df_filtered$total_green)
hist(sqrt(df_filtered$shannon)) # best
hist(df_filtered$simpsons) # best
hist(sqrt(df_filtered$simpsons))

# alternate between shannon/simpsons to explore best (simpsons better)
gam_model3 <- gam(imdd_weight_av ~ s(total_green) + s(log(total_blue)) + s(simpsons) + s(lat), data = df_filtered, family = gaussian())
summary(gam_model3)
par(mfrow = c(2,2))
plot(gam_model3, all.terms = TRUE)
AIC(gam_model3) # 1495
gam.check(gam_model3)

gam_model4 <- gam(imdd_weight_av ~ s(total_green) + s(log(total_blue)) + s(shannon) + s(lat), data = df_filtered, family = gaussian())
summary(gam_model4)
par(mfrow = c(2,2))
AIC(gam_model4) #1520


# Thin plate # BEST MODEL
gam_model3_tp <- gam(imdd_weight_av ~ s(total_green, k=9, bs="tp", fx=FALSE) + 
                                      s(log(total_blue), bs="tp", k=9, fx=FALSE) + 
                                      s(simpsons, k=9, bs="tp", fx=FALSE) +
                                      s(lat, k=9, bs="tp", fx=FALSE),
                  data = df_filtered, family = gaussian())
summary(gam_model3_tp)
par(mfrow = c(2,2))
plot(gam_model3_tp)
AIC(gam_model3_tp)

# method reml = 1595
# GCV = 1588

# Cubic regression
gam_model3_cr <- gam(imdd_weight_av ~ s(total_green, k=, bs="cr", fx=TRUE) + 
                                      s(log(total_blue), bs="cr", k=9, fx=TRUE) + 
                                      s(simpsons, k=9, bs="cr", fx=TRUE) +
                                      s(lat, k=9, bs="cr", fx=FALSE),
                                      data = df_filtered, family = gaussian())
summary(gam_model3_cr)
plot(gam_model3_cr)
AIC(gam_model3_cr) # 1609

# B splines
gam_model3_bs <- gam(imdd_weight_av ~ s(total_green, k=9, bs="bs", fx=FALSE) + 
                                      s(log(total_blue), bs="bs", k=9, fx=FALSE) + 
                                      s(simpsons, k=9, bs="bs", fx=FALSE) +
                                      s(lat, k=9, bs="bs", fx=FALSE),
                  data = df_filtered, family = gaussian())
summary(gam_model3_bs)
par(mfrow = c(1,3))
plot(gam_model3_bs)
AIC(gam_model3_bs) # 1588

```

```{r alternative model}
# alternative model with shannon (does not fit as well)
gam_model3_alt_tp <- gam(imdd_weight_av ~ s(total_green, k=9, bs="tp", fx=FALSE) + 
                                      s(log(total_blue), bs="tp", k=9, fx=FALSE) + 
                                      s(shannon, k=9, bs="tp", fx=FALSE) +
                                      s(lat, k=9, bs="tp", fx=FALSE),
                  data = df_filtered, family = gaussian())
summary(gam_model3_alt_tp)
par(mfrow = c(2,2))
plot(gam_model3_alt_tp)
AIC(gam_model3_alt_tp)
```


```{r gam model export}
library(mgcv)

png(filename = "GAM_Plots.png", width = 3000, height = 3000, res = 300)  # Adjust size as needed

# Set up the plotting parameters to affect all plots
par(mfrow = c(2, 2),    # Arrange the plots in one row and three columns
    cex.main = 1.3,     # Change the main title text size
    cex.lab = 1.3,      # Change the axis labels text size
    cex.axis = 1.5,     # Change the axis ticks text size
    mgp = c(3, 1, 0),   # Set the margins for the title, labels, and axis line
    mar = c(5, 4, 4, 2))# Set the margins around the plot

# Assuming 'gam_model3' has at least three smooth terms, plot each separately
plot(gam_model3_tp, select = 1, main = "", xlab = "Green Space (%)", ylab = "Effect")
plot(gam_model3_tp, select = 2, main = "", xlab = "Blue Space (%)", ylab = "")
plot(gam_model3_tp, select = 3, main = "", xlab = "Simpson's Diversity", ylab = "")
plot(gam_model3_tp, select = 4, main = "", xlab = "Latitude", ylab = "")

# Close the PNG device
dev.off()

# Reset par to default settings after plotting
par(mfrow = c(1, 1), cex.main = 1, cex.lab = 1, cex.axis = 1, mgp = c(2, 1, 0), mar = c(5, 4, 4, 2))

```


```{r gam model 3 check}
par(mfrow = c(2, 3)) # Adjust layout to fit all plots
plot(gam_model3_tp, pages = 1)
par(mfrow = c(1,1)) # Reset layout

# Plot residuals
plot(gam_model3_tp$residuals)
AIC(gam_model3_tp) #1588
BIC(gam_model3_tp) #1641
summary(gam_model3_tp)
par(mfrow = c(2,2))
gam.check(gam_model3_tp)

png("gam_check_output.png", width = 800, height = 800)

par(mfrow = c(2, 2),            # 2x2 layout
    cex.main = 1.5,             # Title size
    cex.lab = 1.4,              # Axis labels
    cex.axis = 1.3,             # Axis tick labels
    cex = 1.3)                  # General text (e.g., point labels)

gam.check(gam_model3_tp)

dev.off()

```

```{r gam model sensitivity check}
# Sensitivity analysis by varying the smoothing parameter 

# Model 1: Lower k value (k=7)
model1 <- gam(imdd_weight_av ~ s(total_green, k=7) + s(log(total_blue), k=7) + s(simpsons, k=7), data = df_filtered)
summary(model1)

# Model 2: Higher k value (k=15) (k can be changed)
model2 <- gam(imdd_weight_av ~ s(total_green, k=15) + s(log(total_blue), k=15) + s(simpsons, k=15), data = df_filtered)
summary(model2)

# Compare the models
summary(model1)
summary(model2)

# Generate predictions for both models
predictions1 <- predict(model1, newdata = df_3)
predictions2 <- predict(model2, newdata = df_3)

# Plot to compare predictions
par(mfrow = c(1, 1))

png("k7_vs_k15_predictions.png", width = 800, height = 800, res = 150)

plot(predictions1, predictions2, 
     main = "k = 7 vs k = 15",
     xlab = "Predictions with k = 7", 
     ylab = "Predictions with k = 15",
     cex.lab = 1.4,       # axis label size
     cex.main = 1.5,      # title size
     cex.axis = 1.3,      # tick label size
     pch = 19, col = "black", size = 3)  # point style and color

abline(0, 1, col = "red", lwd = 2)

dev.off()


# Compare the models
summary(model1)
summary(model2)

```

```{r model cross validation}
# 
set.seed(123)

# Number of folds for cross-validation
n_folds <- 10

# Shuffle and create indices for the folds
fold_indices <- sample(cut(seq(1, nrow(df_filtered)), breaks=n_folds, labels=FALSE))

# Initialize an empty list
cv_results <- vector("list", n_folds)

# Cross-validation
for (i in 1:n_folds) {
  # Define training and test sets based on fold indices
  training_set <- df_filtered[fold_indices != i,]
  test_set <- df_filtered[fold_indices == i,]
  
  # Fit the GAM model using thin plate splines
  gam_model <- gam(imdd_weight_av ~ s(total_green, k=9, bs="tp", fx=FALSE) + 
                                      s(log(total_blue), bs="tp", k=9, fx=FALSE) + 
                                      s(simpsons, k=9, bs="tp", fx=FALSE) +
                                      s(lat, k=9, bs="tp", fx=FALSE), 
                   data = training_set, family = gaussian())
  
  # Summary
  summary(gam_model)# 
  
  # Predict on the test set
  predictions <- predict(gam_model, newdata=test_set)
  
  # Calculate and store the RMSE for this fold
  cv_results[[i]] <- sqrt(mean((test_set$imdd_weight_av - predictions)^2))
}

# Calculate the average RMSE across all folds
average_rmse <- mean(unlist(cv_results))
print(paste("Average RMSE from CV:", average_rmse))

```
